### 2025-05-05 manuscript/manuscript.Rmd
Adding figure 1 heatmap

```{r setup}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(ggh4x)
library(patchwork)
library(ggtext)
library(minpack.lm)
```

@@

>>>>

```{r setup}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(ggh4x)
library(patchwork)
library(ggtext)
library(minpack.lm)

parametadata <- read_csv("metadata/parametadata.csv") %>%
  # filter(IS_type=="noIS") %>%
  filter(polarity=="pos") %>%
  filter(samp_type=="Smp") %>%
  mutate(startime=factor(startime, levels=c("Morn", "Eve"))) %>%
  mutate(depth=factor(depth, levels=c("Surf", "Deep"))) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"))) %>%
  arrange(amendment, depth, timepoint_num, startime) %>%
  mutate(timepoint=fct_inorder(timepoint))
peak_areas <- read_csv("targeted/all_peak_areas_w_isos.csv") %>%
  left_join(parametadata %>% distinct(filename, shortname))
all_stans <- read_csv("metadata/all_stans.csv")
maa_names <- c("Also Palythene/Usujirene?", "Shinorine?", "Mycosporine-glycine?", 
               "Palythine?", "Another Palythene/Usujirene?", "Porphyra-334?", 
               "Also Porphyra-334?", "Asterina-330?", "Palythene/Usujirene?",
               "Mycosporine-2-glycine?")
nucleo_names <- c("5-Methylcytosine", "Adenosine", "Cytidine", "Adenine", "Guanosine",
                  "Cytosine", "Hypoxanthine/Allopurinol", "Guanine", "Deoxycytidine?",
                  "Deoxyadenosine", "Deoxyguanosine?")
betaine_names <- c("beta-Alaninebetaine", "Glycine betaine", "Homarine", "Trigonelline",
                   "Choline", "Glycerophosphocholine", "Homoserine betaine?", 
                   "Threonine betaine?", "Proline betaine",
                   "(3-Carboxypropyl)trimethylammonium", "O-Acetylcarnitine", 
                   "Betonicine", "Acetylcholine", "Phosphocholine")
non_n_names <- c("Dimethylsulfonioacetate", "Dimethylsulfoniopropionate",
                 "Arsenobetaine", "Dimethylsulfoxonium propionate?", "Gonyol",
                 "2-O-alpha-D-Glucosylglycerol")
aa_names <- c("Glutamine", "Aspartic acid", "Glutamic acid/beta-Glutamic acid",
              "Asparagine", "Alanine", "Methionine", "Arginine", "Proline",
              "Lysine", "Threonine/Homoserine", "Serine", "Glycine", "Histidine",
              "Leucine/Isoleucine", "Valine", "Arginine")
bad_cmpds <- c("Melamine", "Uridine", "Itaconic acid?", "Inosine", "Dexpanthenol", 
               "Choline sulfate", "5-Hydroxyectoine", "L-Tyrosine")
samp_areas <- parametadata %>%
  filter(samp_type=="Smp") %>%
  distinct(shortname, depth) %>%
  left_join(peak_areas) %>%
  mutate(shortname=fct_inorder(shortname)) %>%
  filter(!compound_name%in%bad_cmpds) %>%
  filter(!(depth=="Deep" & compound_name%in%maa_names)) %>%
  select(-depth)
labeled_fracs <- samp_areas %>%
  mutate(labeled_frac=1-area/sum(area), .by = c("shortname", "compound_name")) %>%
  select(compound_name, iso_name, shortname, labeled_frac) %>%
  filter(str_detect(iso_name, "13C0, 15N0")) %>%
  select(-iso_name)
n_labeled_fracs <- samp_areas %>%
  filter(str_detect(iso_name, "13C0")) %>%
  filter(!str_detect(iso_name, "15N0") | str_detect(iso_name, "13C0, 15N0")) %>%
  mutate(labeled_frac=1-area/sum(area), .by = c("shortname", "compound_name")) %>%
  select(compound_name, iso_name, shortname, labeled_frac) %>%
  filter(str_detect(iso_name, "13C0, 15N0")) %>%
  select(-iso_name)
log_coef_df <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, amendment, depth, timepoint_num)) %>%
  nest(data=c(shortname, timepoint_num, labeled_frac)) %>%
  mutate(log_fit=map(data, ~glm(.x$labeled_frac~.x$timepoint_num, family="quasibinomial")),
         coefs=map(log_fit, function(x)broom::tidy(x))) %>%
  unnest(coefs) %>%
  select(compound_name, amendment, depth, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(intercept=`(Intercept)`, slope=`.x$timepoint_num`) %>%
  mutate(half_max=-intercept/slope, scale=1/slope) %>%
  mutate(frac_at_T24 = 1/(1+exp(-(intercept+slope*24))))
```

@@

```{r figure 1 }
class_levels <- c("Glutamine", "Glutamic acid", "Aspartic acid", "Asparagine", "Other amino acids", "Theanine?", "Citrulline", "Ectoine", "Guanine", "Other nucleic acids", "MAAs", "Betaines", "All others")

labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint, timepoint_num, depth, startime, amendment)) %>%
  filter(amendment=="Amm") %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%class_levels ~ compound_name,
    compound_name%in%maa_names~"MAAs",
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=class_levels)) %>%
  summarise(avg_frac=mean(labeled_frac), .by = c(compound_class, timepoint_num, startime, depth)) %>%
  mutate(color_group=paste(depth, startime)) %>%
  ggplot(aes(x=timepoint_num, y=avg_frac, color=color_group, fill=color_group)) +
  geom_smooth(method = "glm", formula = y~x, method.args = list(family = "quasibinomial"),
              fill="grey80") +
  geom_point(pch=21, color="black") +
  facet_wrap(~compound_class, ncol=5) +
  scale_y_continuous(labels = scales::label_percent(), breaks = c(0, 0.5, 1)) +
  scale_x_continuous(breaks = c(0.3, 1, 3, 10, 26), labels = c(0.3, 1, 3, 10, 26)) +
  scale_color_manual(breaks = c("Surf Morn", "Surf Eve", "Deep Morn", "Deep Eve"),
                     values = c("#FFD53D", "#DBA20A", "#256A76", "#003743")) +
  scale_fill_manual(breaks = c("Surf Morn", "Surf Eve", "Deep Morn", "Deep Eve"),
                     values = c("#FFD53D", "#DBA20A", "#256A76", "#003743")) +
  labs(x="Hours after 15N added", y="Fraction of metabolite labeled", color=NULL) +
  theme_bw() +
  guides(color = guide_legend(nrow = 2, override.aes = list(linewidth=2)),
         fill = "none") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.8, 1/6),
        legend.location = "plot")

labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(amendment=="Amm") %>%
  mutate(depth=factor(depth, levels=c("Surf", "Deep"), labels=c("Surface seawater (25 meters)", "Seawater from below DCM (175m)"))) %>%
  mutate(startime=factor(startime, levels=c("Morn", "Eve"), labels=c("6AM start", "6PM start"))) %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine") ~ compound_name,
    compound_name%in%maa_names~"MAAs",
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=class_levels)) %>%
  summarise(avg_frac=mean(labeled_frac), .by = c(compound_class, timepoint, startime, depth, tripl)) %>%
  ggplot() +
  geom_raster(aes(x=timepoint, y=tripl, fill=avg_frac)) +
  geom_vline(xintercept = c(1.5:4.5), color="grey80") +
  facet_nested(compound_class~depth+startime, scales = "free_y", switch="y") +
  scale_fill_gradientn(colours = viridis::viridis(10, alpha = seq(0, 1, length.out=10), direction = -1),
                       limits=c(0, 1), breaks=c(0, 0.25, 0.5, 0.75, 1),
                       labels=c("Fully unlabeled", "", "50% labeled", "", "Fully labeled")) +
  scale_x_discrete(expand = expansion()) +
  scale_y_discrete(expand = expansion()) +
  labs(x=NULL, y=NULL, fill=NULL) +
  theme_bw() +
  theme(panel.spacing.x=unit(0.5, "lines"), 
        panel.spacing.y = unit(0, "lines"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle=0, hjust = 1),
        strip.background.y = element_rect(fill=NA, color=NA),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.key.width = unit(1, "null")) +
  guides(
    fill=guide_colorbar(theme=theme(
      legend.frame = element_rect(color = "black"),
      legend.ticks = element_line(color="black"),
      legend.text = ggtext::element_markdown(hjust=c(0, 0, 0.5, 1, 1)))
    )
  )
```
